Incremental Forward Reasoning for White-Box~Proof~Search

### 2025-02-19 paper.tex
Fix figure indentation, spacing
Fix figure spacing
Rewrite a para in intro

ACL2, PVS and Aesop additionally provide special support for \emph{forward reasoning}, also known as \emph{forward chaining}, in the form of \emph{forward rules}.
Given, for example, the transitivity lemma $\All{x~y~z}{x < y → y < z → x < z}$, a forward rule run on the goal $h₁ : a < b,\, h₂ : b < c ⊢ A$ produces the single subgoal $h₁ : a < b,\, h₂ : b < c,\, h₃ : a < c ⊢ A$.
Hence forward rules can be used to establish additional facts that may be relevant for other rules.
For example, the Aeneas project~\cite{Aeneas} uses Aesop's forward rules \jcom{ensure that this is accurate} in its proof automation to establish facts that are later used by an arithmetic decision procedure.

However, Aesop's support for forward rules was, until recently, quite basic.
In particular, each forward rule was tried on each goal separately (subject to pre-filtering by a discrimination tree~\cite{DiscriminationTrees} index).
This is wasteful: while in principle any rule can modify the context in any way, in practice many rules produce subgoals with a similar---or even the same---context as the initial goal.
For example, the ∧-introduction rule (and, in general, any backwards rule) does not affect the goal's context.
Still, all forward rules applied to the initial goal $Γ ⊢ A ∧ B$ would be tried again on the subgoals $Γ ⊢ A$ and $Γ ⊢ B$, with predictably little effect.

In this paper, we present an alternative way to apply forward rules that addresses this issue.
The central idea is to remember, for each goal $Γ ⊢ A$, which forward rules can be partially applied to which sets of hypotheses.
When a non-forward rule then generates subgoals $Δ₁ ⊢ A₁$, \dots, $Δₙ ⊢ Aₙ$, the new contexts $Δᵢ$ are likely similar to $Γ$.
Hence it is generally cheaper (and often much cheaper) to start with the partial matches for $Γ$ and update them for each of the $Δᵢ$.
The update first deletes partial matches that include hypotheses no longer present in $Δᵢ$.
Afterwards it adds new partial matches for the hypotheses that are new in $Δᵢ$, extending the existing partial matches for a rule if the new hypothesis is compatible with them.
Whenever this process produces a full match---i.e., all premises of a forward rule are satisfied by a set of hypothesis in $Δᵢ$---the rule is applied and adds a corresponding new hypothesis to $Δᵢ$.
Our main contribution is a data structure, the \emph{forward state}, that organises the partial matches in such a way that the update can be done efficiently.

In addition to this \emph{incrementality}, the forward state satisfies two other desiderata:
\begin{itemize}
  \item It is \emph{persistent}, i.e.\ forward states for parent and child goals share much of their structure.
        This is desirable since Aesop supports search strategies other than depth-first search and must therefore keep all goals encountered during the search, along with their forward states, in memory.
  \item It reduces the number of \emph{unifications} performed when matching hypotheses against rule premises.
        In dependent type theory, unification happens up to definitional equality, which makes it quite expensive.
        We therefore want to avoid redundant unifications.
\end{itemize}

@@

\begin{center}
  \begin{minipage}{0.44\textwidth}
    \begin{center}
    \begin{equation}\label{eq:example1:state1}
      H_y\colon
      \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, outer sep=auto, level distance=4em, scale=0.5]
            \node at (0,0) {2}
              child {node {$b$}
                child {node {$\{h₁\}$}}};
      \end{tikzpicture}
      M_y\colon
      \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, outer sep=auto, level distance=4em, scale=0.5]
            \node at (0, 0) {1}
              child {node {$c$}
                child {node {$\{[h₁]\}$}}};
      \end{tikzpicture}
    \end{equation}
  \end{center}
  \end{minipage}
  \begin{minipage}{0.55\textwidth}
    \begin{center}
    \begin{equation}\label{eq:example1:state2}
      H_y\colon
      \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, level distance=4em, scale=0.5]
            \node at (0,0) {2}
            child {node[xshift=-2] {$b$}
              child {node {$\{h₁\}$}}}
            child {node[xshift=2] {$a$}
              child {node {$\{h₂\}$}}};
      \end{tikzpicture}
      M_y\colon
      \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, outer sep=auto, level distance=4em, scale=0.5]
            \node at (-1, 0) {1}
            child {node[xshift=-3] {$c$}
              child {node {$\{[h₁]\}$}}}
            child {node[xshift=1] {$b$}
              child {node {$\{[h₂]\}$}}};
      \end{tikzpicture}
    \end{equation}
    \end{center}
  \end{minipage}
\end{center}

>>>>

ACL2, PVS and Aesop additionally provide special support for \emph{forward reasoning}, also known as \emph{forward chaining}, in the form of \emph{forward rules}.
Given, for example, the transitivity lemma $\All{x~y~z}{x < y → y < z → x < z}$, a forward rule run on the goal $h₁ : a < b,\, h₂ : b < c ⊢ A$ produces the single subgoal $h₁ : a < b,\, h₂ : b < c,\, h₃ : a < c ⊢ A$.
Hence forward rules can be used to establish additional facts that may be relevant for other rules.
For example, the Aeneas project~\cite{Aeneas} uses Aesop's forward rules in its proof automation to establish facts that are later used by an arithmetic decision procedure.

Until recently, Aesop used what we call the \emph{naive algorithm} to apply forward rules.
When a forward rule was run on a goal, the rule would search the goal's context for all combinations of hypotheses satisfying its premises.
This method is simple, but involves much repeated work because in practice, contexts encountered during the search tend to be similar.
For example, the ∧-introduction rule, when run on $Γ ⊢ A ∧ B$, produces subgoals $Γ ⊢ A$ and $Γ ⊢ B$, so the naive algorithm runs the same forward rules on the same context three times.
More generally, any backwards rule leaves the context unchanged, and many rules that change the context do so only in limited ways.

In this paper, we present an alternative way to apply forward rules that addresses this issue.
The central idea is to remember, for each goal $Γ ⊢ A$, which forward rules can be partially applied to which sets of hypotheses.
When a non-forward rule then generates subgoals $Δ₁ ⊢ A₁$, \dots, $Δₙ ⊢ Aₙ$, the new contexts $Δᵢ$ are likely similar to $Γ$.
Hence it is generally cheaper (and often much cheaper) to start with the partial matches for $Γ$ and update them for each of the $Δᵢ$.
The update first deletes partial matches that include hypotheses no longer present in $Δᵢ$.
Afterwards it adds new partial matches for the hypotheses that are new in $Δᵢ$, extending the existing partial matches for a rule if the new hypothesis is compatible with them.
Whenever this process produces a full match---i.e., all premises of a forward rule are satisfied by a set of hypothesis in $Δᵢ$---the rule is applied and adds a corresponding new hypothesis to $Δᵢ$.
Our main contribution is a data structure, the \emph{forward state}, that organises the partial matches in such a way that the update can be done efficiently.
In addition to this \emph{incrementality}, the forward state satisfies two other desiderata:
\begin{itemize}
  \item It is \emph{persistent}, i.e.\ forward states for parent and child goals share much of their structure.
        This is desirable since Aesop supports search strategies other than depth-first search and must therefore keep all goals encountered during the search, along with their forward states, in memory.
  \item It reduces the number of \emph{unifications} performed when matching hypotheses against rule premises.
        In dependent type theory, unification happens up to definitional equality, which makes it quite expensive.
        We therefore want to avoid redundant unifications.
\end{itemize}

@@

\begin{center}
  \begin{minipage}{0.43\textwidth}
    \begin{center}
      \begin{align}\label{eq:example1:state1}
        H_y\colon
        \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, outer sep=auto, level distance=4em, scale=0.5]
          \node at (0,0) {2}
            child {node {$b$}
              child {node {$\{h₁\}$}}};
        \end{tikzpicture}
        && M_y\colon
        \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, outer sep=auto, level distance=4em, scale=0.5]
          \node at (0, 0) {1}
            child {node {$c$}
              child {node {$\{[h₁]\}$}}};
        \end{tikzpicture}
      \end{align}
    \end{center}
  \end{minipage}
  \begin{minipage}{0.56\textwidth}
    \begin{center}
      \begin{align}\label{eq:example1:state2}
        H_y\colon
        \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, level distance=4em, scale=0.5]
          \node at (0,0) {2}
            child {node[xshift=-2] {$b$}
              child {node {$\{h₁\}$}}}
            child {node[xshift=2] {$a$}
              child {node {$\{h₂\}$}}};
        \end{tikzpicture}
        && M_y\colon
        \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, outer sep=auto, level distance=4em, scale=0.5]
          \node at (-1, 0) {1}
            child {node[xshift=-4] {$c$}
              child {node {$\{[h₁]\}$}}}
            child {node[xshift=4] {$b$}
              child {node {$\{[h₂]\}$}}};
        \end{tikzpicture}
      \end{align}
    \end{center}
  \end{minipage}
\end{center}
